### 2019-03-20 paper/sex_selection_spacing-ver2.tex
Restructuring Results section
Still missing distributional results and regional results.

The central question addressed here is the extent to which spacing patterns 
significantly changed as prenatal sex determination became available.
Specifically, did the spacing after girls increase relative to the spacing
after boys?
The underlying idea is that in the absence of sex selection son 
preference leads to shorter spacing after the birth of a girl than after the 
birth of a boy, whereas son preference increases the spacing after the birth 
of a girl relative to after the birth of a boy when prenatal sex selection is 
available.
I introduce a new method that simultaneously accounts for spacing
between births and the potential use of sex selection. 
I apply the method to over four decades of data from India's NFHS.
The results show two very different approaches to son preference.

At one extreme, women without education mostly follow the standard 
pattern of shorter spacing when a woman does not have the desired number of 
sons.
There is also limited evidence of biased sex ratios and the increases in
spacing that should follow from the use of sex selection.
 In the situations where there are elevated sex ratios for women without
education, the spacing patterns are close to those with more regular sex
ratios. 
Hence, it is difficult to conclude that the changes in sex
ratios indicate sex selection rather than normal variation or problems
with recall errors.

At the other extreme, women with eight or more years of education have
experienced an almost complete reversal of the traditional spacing patterns.
Rather than having the next birth sooner as they mostly did before 
sex selection became available, women with either no or one son now 
have substantially longer spacing than if they have two or more sons.
As shown by the increasingly significant deviations in the sex ratio
from the natural sex ratio, these changes are likely due to increased
use of sex selection rather than changes in son preference. 
Banning prenatal sex determination has not been able to
reverse the use, although it is possible that it has slowed the rate 
of increase.

Over the last 40 years, there has been a general increase in median
spacing for all education groups, but especially for the most 
educated women.
The exception is for the spacing between marriage and first birth
where there are two counteracting forces: women have become less likely 
to conceive before marriage, and there has been a substantial compression 
in the variation of spacing.
The result is that median spacing from marriage to first birth
has changed little or even declined for the best-educated women.
Both the decrease for the first spell and the increase for higher-order
spells are consistent with increased access to contraceptives over
time \citep{Yeakey2009}.

Although it is not possible to directly observe whether sex selection
causes the unequal sex ratio and longer spacing the results are
consistent with a general increase in sex selection over time. 
The apparent increase in the use of sex selection is mainly found among
women with more education and more so in urban than in rural areas,
although the differences between urban and rural areas are relatively
minor. 
Compared to previous research one result stands out. 
It appears that the use of sex selection is spreading to women with lower 
levels of education.
Even women with one to seven years of education show unequal sex ratios 
and longer spacing for their higher-order spells.

The results here point to two important, interconnected, questions 
for future research:
What is the connection between falling fertility and the 
use of sex selection and is the use of sex selection increasing 
or decreasing over time?
It is clear that fertility has been declining over time in India,
although it is important to note that the decline might be
overstated precisely because of increased use of sex selection and
the resulting longer spacing between births makes it less likely 
that we will observe births to women, even though they eventually
will happen.
The fact that biased sex ratio occurs at lower-order spells over
time and is starting to show up for women with less education 
all indicate that lower fertility and use of sex selection are
related, but a more detailed analysis is needed.
Furthermore, although examining spells separately, as done here, is a useful
first step in understanding how sex selection changes over time,
it does not lend itself easily to examine the degree to which
sex selection changes over time.
For both, a combined model that incorporates the full fertility
history of women would be more appropriate.

The results here also lead to a set of broader questions that
future research should address.
First, to what extent are the improvements in health for girls
relative to boys the result of selection, the longer spacing between
births, or changing son preference?
There is some evidence that sample selection can make it appear that girls are healthier, 
even though the underlying cause is a combination of sex selection and higher mortality 
together with recall error \citep{Portner2018a}. 
Understanding what role these factors play in better health for
girls is especially important when evaluating policies that 
aim to directly limit the use of sex selection rather than 
changing preferences or incentives.
If the better health outcomes for girls are, for example, an unintended 
side-effect of the longer spacing that arises from sex-selective 
abortions, then an effective ban on sex selection may, at least
temporarily, worsen health outcomes for girls.

[move to conclusion?]
We know that health outcomes for girls appear to improve in the presence of sex 
selection \citep{Lin2014,Hu2015}.
It is possible that these improvements are an unintended side-effect of the longer spacing 
that arises from sex-selective abortions, rather than because the smaller number of girls 
makes them more valued as is often assumed.
A first step

Second, we need to understand how female labor force participation
interacts with the use of sex selection.
Increased autonomy for women, arising, for example, from better opportunities
for working outside the home, has been suggested as a way to increase
women's status and thereby lower the use of sex selection \citep{Das-Gupta2016}.
This may, however, be a double-edged sword.
On the one hand, it is a clear benefit to the women who gain  
bargaining power, and it increases the cost of repeated 
sex-selective abortions because the increased duration between births
would cause a stronger disruption in labor market participation.
On the other hand, it may further lower desired fertility, and 
that may, everything else equal, lead to higher use of sex selection.
Understanding the trade-off between long-term benefits from improvements in
women's labor force participation and short-term costs from potential 
increases in sex selection is of paramount importance.

>>>>

The central question addressed here is the extent to which spacing patterns significantly 
changed as prenatal sex determination became available.
The underlying idea is that in the absence of sex selection son preference leads to shorter 
spacing after the birth of a girl than after the birth of a boy, whereas when prenatal sex 
selection is available son preference increases the spacing after the birth of a girl 
relative to after the birth of a boy.
I introduce an empirical method that simultaneously accounts for spacing between births 
and the potential use of sex selection. 
I apply the method to over four decades of data from India's NFHS.

The results show three major, interrelated trends over the last 40 years.
First, fertility has declined for all groups, as shown by the parity progression
probabilities, with a larger decline the more educated the woman.
Across all groups, the likelihood of having an additional child in each spell depends 
strongly on how many sons you already have, with women with none or only one son having
the highest probability of a birth.
Second, there has been a general increase in the length of average birth 
intervals for all education groups.
The increase in average birth interval is larger the higher the spell and the higher the 
education group.
Finally, as shown by the increasing sex ratios for women with no or only one son there
has been a substantial increase in the use of sex selection over time.
As for fertility and birth intervals, this increase is larger the higher the spell and
with higher education.

The results also show two very different approaches to son preference with important
impacts on spacing patterns.
At one extreme, rural women without education mostly follow the standard pattern of 
shorter spacing when a woman does not have the desired number of sons.
The strength of this pattern does, however, depend critically on spell number.
For the second spell, although the average interval after a first-born girl is shorter
than the average interval after a first-born boy, the difference is small.
The differences in spacing is larger for the third spell and largest for the fourth.
For the the fourth spell, the divergence in intervals across sex compositions have 
increased over time, with the largest increases in average intervals for women with 
two or three boys.
There has also been an increase in average spacing for women with only girls, and there
is some limited evidence of biased sex ratios, which suggests that this increases in
spacing follow from the use of sex selection.

At the other extreme, urban women with eight or more years of education have experienced 
an almost complete reversal of the traditional spacing patterns as a result of falling
fertility and the associated extensive use of sex selection.
Rather than having the next birth sooner as they mostly did before sex selection became 
available, women with either no or one son now have substantially longer spacing than if 
they have two or more sons.
As shown by the increasingly significant deviations in the sex ratio from the natural sex 
ratio, these changes are likely due to increased use of sex selection rather than sudden
reversal in son preference.%
\footnote{
Banning prenatal sex determination has not been able to reverse the use, although it is 
possible that it has slowed the rate of increase.
}
Only the second spell has not seen a reversal despite a substantial increased use of
sex selection, mainly because the average spacing with a first-born boy has increased 
independently by more than nine months.

TK Distribution results

TK Regional differences

The results here lead to a set of broader questions that future research should address.
First, what is behind the substantial increases in spacing even for women who are unlikely 
to use sex selection, and, related, why does spacing appear to increase with declining 
parity progression probabilities?
The distributional results provide an initial indication that most of the increases in 
average birth intervals come from the upper end of the birth intervals rather an generic
increase across the distribution.
It is, however, unclear why we are seeing relatively more very long birth intervals over
time, and the longer average spacing seem to run counter to a general reduction in the 
length of women's reproductive spans found in earlier research \citep{Padmadas2004}.
Understanding whether the longer average spacing is by choice at the beginning of the 
spell, because of unplanned pregnancies, or because of changes in economic circumstances 
that allow families to have an additional child later has important implications for both 
policy design and outcomes for mother and children.

Second, to what extent are the improvements in health for girls relative to boys the result 
of selection, the longer spacing between births, or changing son preference?
We know that health outcomes for girls appear to improve in the presence of sex selection 
\citep{Lin2014,Hu2015}.
Furthermore, there is some evidence that sample selection can make it appear that girls 
are healthier, even though the underlying cause is a combination of sex selection and 
higher mortality together with recall error \citep{Portner2018a}. 
Understanding what role these factors play in better health for girls is especially 
important when evaluating policies that aim to directly limit the use of sex selection 
rather than changing preferences or incentives.
If the better health outcomes for girls are, for example, an unintended side-effect of the 
longer spacing that arises from sex-selective abortions, rather than because the smaller 
number of girls makes them more valued as is often assumed, then an effective ban on sex 
selection may, at least temporarily, worsen health outcomes for girls.

Finally, we need to understand how female labor force participation interacts with the use 
of sex selection.
Increased autonomy for women, arising, for example, from better opportunities for working 
outside the home, has been suggested as a way to increase women's status and thereby lower 
the use of sex selection \citep{Das-Gupta2016}.
This may, however, be a double-edged sword.
On the one hand, it is a clear benefit to the women who gain bargaining power, and it 
increases the cost of repeated sex-selective abortions because the increased duration 
between births would cause a stronger disruption in labor market participation.
On the other hand, it may further lower desired fertility, and that may, everything else 
equal, lead to higher use of sex selection.
Understanding the trade-off between long-term benefits from improvements in women's labor 
force participation and short-term costs from potential increases in sex selection is of 
paramount importance.
